Higher-Order unification for free!

### 2024-04-12 paper/main.tex
sec 2.1 with spellchecker

Even if we encountered the problem working on CIC we devise
a minimal setting to ease its study. In this setting we have
a \Fo{} language (for first order) with a rich equational
theory and a \Ho{} meta language with a simpler one.

@@

In order to reason about unification we provide a description of the
\Fo{} and \Ho languages where unification variables
are first class terms, i.e. they have a concrete syntax as
per \cref{code:common-terms}.
% We keep these languages
%minimal, for example, we omit the \elpiIn{all} quantifier of CIC we used
%in the example in Section~\ref{sec:intro} together with the type notation of
%terms carried by the \elpiIn{lam} constructor.
%
{
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{-13pt}
\begin{figure}[b] %[H]
  \begin{tabular}{ll}
  \begin{minipage}{0.21\textwidth}
   \inputrawelpicode{code/fo_tm}
  \end{minipage}
  &
  \begin{minipage}{0.24\textwidth}
   \inputrawelpicode{code/ho_tm}
  \end{minipage}
  \end{tabular}\vspace{4pt}
  \caption{The \Fo{} and \Ho languages}\vspace{0.3em}
  \label{code:common-terms}
  \Description[code:common-terms]{code:common-terms}
\end{figure}
}
Unification variables
in \Fo{} (\elpiIn{fuva} term constructor) have no explicit scope:
the arguments of an higher order variable are given via the \elpiIn{fapp}
constructor. For example the term <<\coqIn{P x}>> is represented as
<<\elpiIn{fapp [fuva N, x]}>>, where \elpiIn{N} is the memory address
of \elpiIn{P} and \elpiIn{x} is a bound variable.\\
In \Ho the representation of <<\coqIn{P x}>> is instead <<\elpiIn{uva N [x]}>>,
since unification variables are higher order and come equipped with an
explicit scope.
%\todo{use faddr for Fo}
%
%faddr and addr are memory addresses, the details are given in bla,
%here they are just unique identifiers for unif variables and their types
%do now allow to mistake one for the other.
%\todo{cleanup}

@@

When we write \Ho terms outside code blocks we follow the
usual $\lambda$-calculus notation, reserving $f, g, a, b$ for constants,
$x, y, z$ for bound variables and $X, Y, Z, F, G, H$ for unification variables.
However we need to
distinguish between the ``application'' of a unification variable
to its scope and the application of a term to a list of arguments.
We write the scope of unification variables in subscript
while we use juxtaposition for regular application.
Here a few examples:\\
\vspace{4pt}
{
\setlength{\tabcolsep}{1em}
\begin{tabular}{ll}
  $f\appsep a$                  & \elpiIn{app [con "f", con "a"]}\\
  $\lambda x.\lambda y.F_{x y}$ & \elpiIn{lam x\ lam y\ uva F [x, y]} \\
  $\lambda x.F_{x} \appsep a$   & \elpiIn{lam x\ app [uva F [x], con "a"]} \\
  $\lambda x.F_{x} \appsep x$   & \elpiIn{lam x\ app [uva F [x], x]} \\
\end{tabular}
}
\vspace{4pt}

\noindent
When it is clear from the context we shall use the same syntax for \Fo{} terms
(although we never subscripts unification variables).\\
We use $s$, $s_1$, \ldots for terms in \Fo{} and $t$, $t_1$ \ldots for
terms in \Ho{}.

>>>>

Even if we encountered the problem working on CIC, we devise
a minimal setting to ease its study. In this setting, we have
a \Fo{} language (for first order) with a rich equational
theory and a \Ho{} meta language with a simpler one.

@@

To reason about unification, we provide a description of the
\Fo{} and \Ho languages where unification variables
are first-class terms, i.e. they have a concrete syntax as
shown in \cref{code:common-terms}.
% We keep these languages
%minimal, for example, we omit the \elpiIn{all} quantifier of CIC we used
%in the example in Section~\ref{sec:intro} together with the type notation of
%terms carried by the \elpiIn{lam} constructor.
%
{
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{-13pt}
\begin{figure}[b] %[H]
  \begin{tabular}{ll}
  \begin{minipage}{0.21\textwidth}
   \inputrawelpicode{code/fo_tm}
  \end{minipage}
  &
  \begin{minipage}{0.24\textwidth}
   \inputrawelpicode{code/ho_tm}
  \end{minipage}
  \end{tabular}\vspace{4pt}
  \caption{The \Fo{} and \Ho languages}\vspace{0.3em}
  \label{code:common-terms}
  \Description[code:common-terms]{code:common-terms}
\end{figure}
}
Unification variables
in \Fo{} (\elpiIn{fuva} term constructor) have no explicit scope:
the arguments of a higher-order variable are given via the \elpiIn{fapp}
constructor. For example the term <<\coqIn{P x}>> is represented as
<<\elpiIn{fapp [fuva N, x]}>>, where \elpiIn{N} is the memory address
of \elpiIn{P} and \elpiIn{x} is a bound variable.\\
In \Ho, the representation of <<\coqIn{P x}>> is instead <<\elpiIn{uva N [x]}>>,
since unification variables are higher-order and come equipped with an
explicit scope.
%\todo{use faddr for Fo}
%
%faddr and addr are memory addresses, the details are given in bla,
%here they are just unique identifiers for unif variables and their types
%do now allow to mistake one for the other.
%\todo{cleanup}

@@

When we write \Ho terms outside code blocks we follow the
usual $\lambda$-calculus notation, reserving $f, g, a, b$ for constants,
$x, y, z$ for bound variables and $X, Y, Z, F, G, H$ for unification variables.
However, we need to
distinguish between the ``application'' of a unification variable
to its scope and the application of a term to a list of arguments.
We write the scope of unification variables in subscript
while we use juxtaposition for regular application.
Here are few examples:\\
\vspace{4pt}
{
\setlength{\tabcolsep}{1em}
\begin{tabular}{ll}
  $f\appsep a$                  & \elpiIn{app [con "f", con "a"]}\\
  $\lambda x.\lambda y.F_{x y}$ & \elpiIn{lam x\ lam y\ uva F [x, y]} \\
  $\lambda x.F_{x} \appsep a$   & \elpiIn{lam x\ app [uva F [x], con "a"]} \\
  $\lambda x.F_{x} \appsep x$   & \elpiIn{lam x\ app [uva F [x], x]} \\
\end{tabular}
}
\vspace{4pt}

\noindent
When it is clear from the context, we shall use the same syntax for \Fo{} terms
(although we never subscripts unification variables).\\
We use $s$, $s_1$, \ldots for terms in \Fo{} and $t$, $t_1$, \ldots for
terms in \Ho{}.
